---
title: "Lab8"
author: "Kelsie Fronheiser"
date: "2/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(janitor)
library(palmerpenguins)

# Packages for cluster analysis:
library(NbClust)
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)
```


### Exploratory visualization
```{r}
# Bill length vs depth
ggplot(penguins) +
  geom_point(aes(x = bill_length_mm, 
                 y = bill_depth_mm, 
                 color = species,
                 shape = sex),
             size = 3,
             alpha = 0.7) +
  scale_color_manual(values = c("orange","cyan4","darkmagenta"))

# Flipper length vs body mass 
ggplot(penguins) +
  geom_point(aes(x = flipper_length_mm, 
                 y = body_mass_g, 
                 color = species,
                 shape = sex),
             size = 3,
             alpha = 0.7) +
  scale_color_manual(values = c("orange","cyan4","darkmagenta"))
```

### Using 'NbClust' to run analysis of possible clusters
```{r}
# How many clusters do you THINK there should be? 
number_est <- NbClust(penguins[3:6], min.nc = 2, max.nc = 10, method = "kmeans")

# Check out the results (just look at the first summary report):
number_est

# By these estimators, 2 is identified as the best number of clusters by the largest number of algorithms (8 / 30) but it still makes sense to sort by species- 3. 
```

### Complete, scaled data
```{r}
# Drop rows where any of the four size measurements are missing
penguins_complete <- penguins %>% 
  drop_na(bill_length_mm, bill_depth_mm, body_mass_g, flipper_length_mm)

# Only keep the columns for the four size measurements, then SCALE them
penguins_scale <- penguins_complete %>% 
  select(ends_with("mm"), body_mass_g) %>% 
  scale() # See ?scale for details on scaling

#penguins_scale
```

## k-means
```{r}
penguins_km <- kmeans(penguins_scale, 3) # kmeans specifying 3 groups to start

penguins_km
```

```{r}
# See what it returns (different elements returned by kmeans function):
penguins_km$size # How many observations assigned to each cluster
penguins_km$cluster # What cluster each observation in penguins_scale is assigned to

# Bind the cluster number to the original data used for clustering, so that we can see what cluster each penguin is assigned to
penguins_cl <- data.frame(penguins_complete, cluster_no = factor(penguins_km$cluster))

# Plot flipper length versus body mass, indicating which cluster each penguin is assigned to (but also showing the actual species):
ggplot(penguins_cl) +
  geom_point(aes(x = flipper_length_mm, 
                 y = body_mass_g, 
                 color = cluster_no,
                 shape = species))

# On your own: do the same, plotting bill dimensions and mapping species & cluster number to the point shape and color aesthetics:

ggplot(penguins_cl) +
  geom_point(aes(x = bill_length_mm,
                 y = bill_depth_mm, 
                 color = cluster_no,
                 shape = species))

```

We see that a lot of gentoos are in Cluster 3, a lot of Adelies are in Cluster 2, and A lot of chinstraps are in Cluster 1...but what are the actual counts? Let's find them: 
```{r}
# Find the counts of each species assigned to each cluster, then pivot_wider() to make it a contingency table:
penguins_cl %>% 
  count(species, cluster_no) %>% 
  pivot_wider(names_from = cluster_no, values_from = n) %>% 
  rename('Cluster 1' = '1', 'Cluster 2' = '2', 'Cluster 3' = '3')
```

Takeaway: *Most* chinstraps in Cluster 1, *most* Adelies in Cluster 2, and *all* Gentoos are in Cluster 3 by k-means clustering. So this actually does a somewhat decent job of splitting up the three species into different clusters, with some overlap in Cluster 1 between Adelies & chinstraps, which is consistent with what we observed in exploratory data visualization.

### Part 2. Cluster analysis: hierarchical - `stats::hclust()` function for agglomerative hierarchical clustering
```{r}
# Get the data
wb_env <- read_csv("wb_env.csv")


#View(wb_env)

# Only keep top 20 greenhouse gas emitters
wb_ghg_20 <- wb_env %>% 
  arrange(-ghg) %>% 
  head(20)
```

```{r}
# Scale the numeric variables (columns 3:7)
wb_scaled <- wb_ghg_20 %>% 
  select(3:7) %>% 
  scale()
  

# Update to add rownames (country name) from wb_ghg_20
rownames(wb_scaled) <- wb_ghg_20$name

# Check the outcome with 
View(wb_scaled)
#see that the row names are now the country name (this is useful for visualizing)
```

Now we have a simplified, scaled version of the numeric variables, with rownames containing the county name. 

### Use `stats::dist()` function to find the Euclidean distance in multivariate space between the different observations

```{r}
# Compute dissimilarity values (Euclidean distances):
euc_distance <- dist(wb_scaled, method = "euclidean")

# Check out the output:
#View(euc_distance)
```

#### Perform hierarchical clustering by complete linkage with `stats::hclust()`

The `stats::hclust()` function performs hierarchical clustering, given a dissimilarity matrix (our matrix of euclidean distances), using a linkage that you specify - complete linkage

```{r}
# Hierarchical clustering (complete linkage)
hc_complete <- hclust(euc_distance, method = "complete" )

# Plot it (base plot):
plot(hc_complete, cex = 0.6, hang = -1)
```

### Single linkage
```{r}
# Hierarchical clustering (single linkage)
hc_single <- hclust(euc_distance, method = "single" )

# Plot it (base plot):
plot(hc_single, cex = 0.6, hang = -1)
```

We see that it is a bit different when we change the linkage! But how different? 

#### Make a tanglegram to compare dendrograms 

Let's make a **tanglegram** to compare clustering by complete and single linkage! We'll use the `dendextend::tanglegram()` function to make it. 

First, we'll convert to class `dendrogram`, then combine them into a list:
```{r}
# Convert to class dendrogram
dend_complete <- as.dendrogram(hc_complete)
dend_simple <- as.dendrogram(hc_single)

# Make a tanglegram
tanglegram(dend_complete, dend_simple)
```

That allows us to compare how things are clustered by the different linkages!

### Ggplot dendrogram 

Here's how you can make your dendrogram with `ggplot` (here, I'll use the complete linkage example stored as `hc_complete`) using `ggdendrogram()`, a `ggplot` wrapper:
```{r}
ggdendrogram(hc_complete, 
             rotate = TRUE) +
  theme_minimal() +
  labs(x = "Country")

#Can customize using usual ggplot tools
```

